\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2009-07-13 papers/jss/article.tex
Updated refs in the jss paper (still have compile problems ...)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@295 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[article]{jss}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Ingmar Visser\\University of Amsterdam \And 
        Maarten Speekenbrink\\University College London }
\title{\pkg{depmixS4}: An \proglang{R}-package for hidden Markov models}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated
\Plaintitle{depmixS4: An R-package for hidden Markov models} %% without formatting
\Shorttitle{depmixS4 Hidden Markov Models} %% a short title (if necessary)

%% an abstract and keywords
	\Abstract{

	\pkg{depmixS4} implements a general framework for definining and
	fitting hidden Markov mixture model in the R programming language
	\citep{R2009}.  This includes standard Markov models,
	latent/hidden Markov models, and latent class and finite mixture
	distribution models.  The models can be fitted on mixed
	multivariate data with multinomial and/or gaussian distributions.
	Parameters can be estimated subject to general linear constraints.
	Parameter estimation is done through an EM algorithm or by a
	direct optimization approach using the \pkg{Rdonlp2} optimization
	routine when contraints are imposed on the parameters.  A number
	of illustrative examples are included.

}

\Keywords{hidden Markov model, dependent mixture model, mixture model, \proglang{R}}

\Plainkeywords{hidden Markov model, dependent mixture model, mixture model,R} %% without formatting

%% at least one keyword must be supplied

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Ingmar Visser\\
Department of Psychology\\
University of Amsterdam,
Roetersstraat 15\\
1018 WB, Amsterdam\\
The Netherlands\\
E-mail: \email{i.visser@uva.nl} \\
URL: \url{http://www.ingmar.org/}
}

@@

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

@@

\section[]{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

@@

\subsection*{Acknowledgements}

Ingmar Visser was supported by an EC Framework 6 grant, project 516542
(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
Economic Learning and Social Evolution (ELSE).  Han van der Maas
provided the speed-accuracy data \cite{Maas2005b} and thereby
neccessitated implementing models with time-dependent covariates.
Brenda Jansen provided the balance scale data set \citep{Jansen2001}
which was the perfect opportunity to test the covariates on the prior
model parameters.  The examples in the help files use both of these
data sets.

@@

The dependent mixture model is then given by the following equations: 
\begin{align}
	S_{t} = A S_{t-1}, t=2, \ldots, T \\
	O_{t} = f(O_{t}|S_{t}),
\end{align}
where $S_{t}$ is a sequence of hidden states, $A$ is a transition
matrix, $O_{t}$ is an (possibly multivariate) observation and $f$ is a
density function for $O_{t}$ conditional on the hidden state $S_{t}$.
In the example data above, $f$ could be a gaussian distribution
function for the response time variable, and a Bernouilli for the
accuracy data.  In the models we are considering here, both the
transition probabilities $A$ and the initial state probabilities $\pi$
may depend on covariates as well as the response distributions $f$. 
See for example \citet{Fruhwirth-Schnatter2006} for an overview of 
hidden Markov models with extensions.

@@

The \pkg{depmixS4} contains another data set which is used to
illustrate the use of covariates on the prior probabilities of models,
in this case a latent class model. The \code{balance} data consists 
of 4 binary items (correct-incorrect) on a balance scale task 
\citet{Siegler198?}. The data form a subset of the data published in 
\citet{Jansen199?}.

@@

In the \code{speed} data example, it may be more appropriate to model
the response times as an exgaus distribution rather than using the
gaussian.  We can do so as follows, first we define our own
exgaus-class extending the response-class and set it to be a generic:
\begin{verbatim}
	setClass("exgaus", contains="response")
	setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
	standardGeneric("exgaus"))
\end{verbatim}

@@

The constructor function is defined as: 
\begin{verbatim}
setMethod("exgaus", 
	signature(y="ANY"), 
	function(y,pstart=NULL,fixed=NULL, ...) {
		y <- matrix(y,length(y))
		x <- matrix(1)
		parameters <- list()
		npar <- 3
		if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
		if(!is.null(pstart)) {
			if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
			parameters$mu <- pstart[1]
			parameters$sigma <- log(pstart[2])
			parameters$nu <- log(pstart[3])
		}
		mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
		mod
	}
)
\end{verbatim}

@@

>>>>

%&LaTeX
\documentclass[article]{jss}

%\usepackage[]{amsmath, amsfonts, amstext, amsthm} 
%\usepackage{amssymb}
%\usepackage[]{graphics} 
%\usepackage{epsfig}
%\usepackage{epstopdf}

%% almost as usual
\author{Ingmar Visser\\University of Amsterdam \And 
        Maarten Speekenbrink\\University College London}

\title{\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Ingmar Visser, Maarten Speekenbrink} %% comma-separated

\Plaintitle{depmixS4: An R-package for hidden Markov models} %% without formatting

\Shorttitle{depmixS4: Hidden Markov Models} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{ 	
	depmixS4 implements a general framework for definining and
	fitting hidden Markov mixture model in the R programming language
	R2009.  This includes standard Markov models,
	latent/hidden Markov models, and latent class and finite mixture
	distribution models.  The models can be fitted on mixed
	multivariate data with multinomial and/or gaussian distributions.
	Parameters can be estimated subject to general linear constraints.
	Parameter estimation is done through an EM algorithm or by a
	direct optimization approach using the Rdonlp2 optimization
	routine when contraints are imposed on the parameters.  A number
	of illustrative examples are included.
}

\Keywords{hidden markov model, dependent mixture model, mixture model}

\Plainkeywords{hidden markov model, dependent mixture model, mixture model} %% without formatting
%% at least one keyword must be supplied

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Ingmar Visser\\
Department of Psychology\\
University of Amsterdam\\
Roetersstraat 15\\
1018 WB, Amsterdam\\
The Netherlands\\
E-mail: \email{i.visser@uva.nl} \\
URL: \url{http://www.ingmar.org/}
}

@@

\newcommand{\vc}{\mathbf}
\newcommand{\mat}{\mathbf}

@@

% Refs to check: Jansen 2002, Dutilh 2009

\section{Introduction}
%% Note: If there is markup in \(sub)section, then it has to be escape as above.

@@

@@

The dependent mixture model is then given by the following equations: 
\begin{align}
	S_{t} = A S_{t-1}, t=2, \ldots, T \\
	O_{t} = f(O_{t}|S_{t}),
\end{align}
where $S_{t}$ is a sequence of hidden states, $A$ is a transition
matrix, $O_{t}$ is an (possibly multivariate) observation and $f$ is a
density function for $O_{t}$ conditional on the hidden state $S_{t}$.
In the example data above, $f$ could be a gaussian distribution
function for the response time variable, and a Bernouilli for the
accuracy data.  In the models we are considering here, both the
transition probabilities $A$ and the initial state probabilities $\pi$
may depend on covariates as well as the response distributions $f$. 
See for example \citet{Fruhwirth2006} for an overview of 
hidden Markov models with extensions.

@@

The \pkg{depmixS4} contains another data set which is used to
illustrate the use of covariates on the prior probabilities of models,
in this case a latent class model. The \code{balance} data consists 
of 4 binary items (correct-incorrect) on a balance scale task 
\citet{Siegler1981}. The data form a subset of the data published in 
\citet{Jansen2002}.

@@

In the \code{speed} data example, it may be more appropriate to model
the response times as an exgaus distribution rather than using the
gaussian.  We can do so as follows, first we define our own
exgaus-class extending the response-class:
\begin{verbatim}
	setClass("exgaus", contains="response")
\end{verbatim}

@@

The constructor function is defined as: 
\begin{verbatim}
setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
	standardGeneric("exgaus"))
setMethod("exgaus", 
	signature(y="ANY"), 
	function(y,pstart=NULL,fixed=NULL, ...) {
		y <- matrix(y,length(y))
		x <- matrix(1)
		parameters <- list()
		npar <- 3
		if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
		if(!is.null(pstart)) {
			if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
			parameters$mu <- pstart[1]
			parameters$sigma <- log(pstart[2])
			parameters$nu <- log(pstart[3])
		}
		mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
		mod
	}
)
\end{verbatim}

@@

\section*{Acknowledgements}

Ingmar Visser was supported by an EC Framework 6 grant, project 516542
(NEST).  Maarten Speekenbrink was supported by the ESRC Centre for
Economic Learning and Social Evolution (ELSE).  Han van der Maas
provided the speed-accuracy data \cite{Dutilh2009} and thereby
neccessitated implementing models with time-dependent covariates.
Brenda Jansen provided the balance scale data set \citep{Jansen2002}
which was the perfect opportunity to test the covariates on the prior
model parameters.  The examples in the help files use both of these
data sets.
